Presentation

### 2017-02-14 paper_project2.Rmd
penultimate commit

## Abstract
Fibromyalgia (FM) is a chronic disorder characterized by widespread, persistent muscle pain throughout the body, often centered at tender points. This pain is generally accompanied by fatigue and sleep problems. The pain associated with fibromyalgia is subjective, and the condition is therefore difficult to measure and treat. This study conducts a randomized clinical trial comparing three treatment therapies: Emotional Awareness and Expression Therapy (EET), Cognitive-behavioral Therapy (CBT), and FM Education (EDU), the control treatment. Using score on the Pain Severity Index (PSI) survey as a measurement for pain, this analysis uses a linear mixed model to determine the difference between the two treatment groups and the control group at two post-treatment time points. Using the results of the model, we show that all three treatments in the study were associated with a significant decrease in pain over time. We then note that the two treatment groups were more effective than the control group at the three-month post-randomization time point, but that the control group then catches up to the treatment groups by the time of the nine-month post-randomization follow-up evaluation.

@@

## Methods
This dataset comes from a two-site, three-arm, allegiance-controlled randomized clinical trial. The dataset contains 230 individual subjects, with 94$\%$ females and 6$\%$ males. 
Subjects at each site were placed into blocks of 18 patients as they entered the study, and then each block was broken into three experimental groups of six, with each group assigned to a different treatment arm. Many baseline variables were measured, including age, education, income, and race. The outcome variable, pain, was measured via a survey at baseline (V2), immediately post-treatment (V3), and six months post-treatment (V4). Each treatment consisted of eight sessions in eight weeks, with each session within the same treatment involving different activities. Many patients did not attend all eight sessions. There were also a number of missing values, with 36 missing pain values out of the total 690 pain measurements across all three time points. 
Prior to fitting a model, the dataset was collapsed on group by taking the mean pain level within each group at each time point, resulting in a total of 120 observations (40 groups with 3 measurements each). Missing pain values were omitted while calculating the mean. Besides collapsing pain measurements by group, no other values were omitted. Collapsing on group loses individual-level data, but serves to reduce the number of random effects and deal with the missing values and missed treatments. Looking on the group level fits naturally with the study design, as each group goes through all eight sessions together and are encouraged to learn and partake in practice exercises with other members of the same group. Subjects who miss sessions will see other familiar group members the following week who did not miss the previous session.

@@

### Model
```{r, message=FALSE, warning=FALSE, results="asis", echo = F}
#group <- lmer(mean_pain ~ time + time:condition + (0 + grpnbr | time), data = group_pain)
#summary(group)
#confint(group, method="boot", nsim=1000)
group <- lme(mean_pain ~ time + time:condition, random = ~ 0 + time | grpnbr, data = group_pain)
#summary(group) # this is the one I'm going with I think; I took mean pain for each group, then two levels.
#stargazer(group, type = "text", report = "vcsp", header = F, digits = 2)
```
\begin{table}[!htbp]
\centering
\begin{tabular}{llll}
\hline
Variable & Value & Std.Error & p-value \\ \hline
timeV3 & -0.22 & 0.22 & 0.30 \\
timeV4 & -0.58 & 0.24 & 0.02 \\
timeV2*conditionCBT & -0.10 & 0.30 & 0.75 \\
timeV3*conditionCBT & -0.56 & 0.32 & 0.09 \\
timeV4*conditionCBT & -0.12 & 0.41 & 0.77 \\
timeV2*conditionEET & -0.13 & 0.29 & 0.66 \\
timeV3*conditionEET & -0.90 & 0.31 & 0.01 \\
timeV4*conditionEET & -0.41 & 0.40 & 0.30
\end{tabular}
\caption{Linear Mixed Model results}
\label{}
 P-values are calcuated using 72 degrees of freedom.
\end{table}

>>>>

## Abstract
Fibromyalgia (FM) is a chronic disorder characterized by widespread, persistent muscle pain throughout the body, often centered at tender points. This pain is generally accompanied by fatigue and sleep problems. The pain associated with fibromyalgia is subjective, and the condition is therefore difficult to measure and treat. This study conducts a randomized clinical trial comparing three treatment therapies: Emotional Awareness and Expression Therapy (EET), Cognitive-behavioral Therapy (CBT), and FM Education (EDU), the control treatment. Using score on the Pain Severity Index (PSI) survey as a measurement for pain, this analysis uses a linear mixed model to determine the difference between the two treatment groups and the control group at two post-treatment time points. Using the results of the model, we show that all three treatments in the study were associated with a significant decrease in pain over time. We then note that the two treatment groups were more effective than the control group at the three-month post-randomization time point, but that the control group then catches up to the treatment groups by the time of the nine-month post-randomization follow-up evaluation. The two treatment groups do not show a loss of effect between the three-month post-randomization time point and the  nine-month post-randomization follow-up evaluation.

@@

## Methods
This dataset comes from a two-site, three-arm, allegiance-controlled randomized clinical trial. The dataset contains 230 individual subjects, with 94$\%$ females and 6$\%$ males. 
Subjects at each site were placed into blocks of 18 patients as they entered the study, and then each block was broken into three experimental groups of six, with each group assigned to a different treatment arm. Many baseline variables were measured, including age, education, income, and race. The outcome variable, pain, was measured via a survey at baseline (V2), immediately post-treatment (V3), and six months post-treatment (V4). Each treatment consisted of eight sessions in eight weeks, with each session within the same treatment involving different activities. Many patients did not attend all eight sessions. There were also a number of missing values, with 36 missing pain values out of the total 690 pain measurements across all three time points.

To assess if there were any differences between subjects in different treatment groups, variables were stratified by treatment and then compared using chi-squared tests if they were categorical, and one-way anova if they were continuous. The normality assumption for the anova tests was visually evaluated using density plots. A p-value cutoff of 0.05 was used to check for significant differences.

Prior to fitting a model, the dataset was collapsed on group by taking the mean pain level within each group at each time point, resulting in a total of 120 observations (40 groups with 3 measurements each). Missing pain values were omitted while calculating the mean. Besides collapsing pain measurements by group, no other values were omitted. Collapsing on group loses individual-level data, but serves to reduce the number of random effects and deal with the missing values and missed treatments. Looking on the group level fits naturally with the study design, as each group goes through all eight sessions together and are encouraged to learn and partake in practice exercises with other members of the same group. Subjects who miss sessions will see other familiar group members the following week who did not miss the previous session.

@@

### Model
```{r, message=FALSE, warning=FALSE, results="asis", echo = F}
#group <- lmer(mean_pain ~ time + time:condition + (0 + grpnbr | time), data = group_pain)
#summary(group)
#confint(group, method="boot", nsim=1000)
#group <- lme(mean_pain ~ time + time:condition, random = ~ time | grpnbr, data = group_pain)
group <- lme(mean_pain ~ time + time:condition, random = ~ 0 + time | grpnbr, data = group_pain)
#summary(group) # this is the one I'm going with I think; I took mean pain for each group, then two levels.
#stargazer(group, type = "text", report = "vcsp", header = F, digits = 2)
```
\begin{table}[!htbp]
\centering
\begin{tabular}{llll}
\hline
Variable & Value & Std.Error & p-value \\ \hline
timeV3 & -0.22 & 0.22 & 0.30 \\
timeV4 & -0.58 & 0.24 & 0.02 \\
timeV2*conditionCBT & -0.10 & 0.30 & 0.75 \\
timeV3*conditionCBT & -0.56 & 0.32 & 0.09 \\
timeV4*conditionCBT & -0.12 & 0.41 & 0.77 \\
timeV2*conditionEET & -0.13 & 0.29 & 0.66 \\
timeV3*conditionEET & -0.90 & 0.31 & 0.01 \\
timeV4*conditionEET & -0.41 & 0.40 & 0.30
\end{tabular}
\caption{Linear Mixed Model results}
\label{}
 P-values are calcuated using 72 degrees of freedom.
\end{table}

### 2017-02-14 presentation.Rmd
penultimate commit

```{r, message=FALSE, warning=FALSE, results="asis", echo = F}
#group <- lmer(mean_pain ~ time + time:condition + (0 + grpnbr | time), data = group_pain)
#summary(group)
#confint(group, method="boot", nsim=1000)
group <- lme(mean_pain ~ time + time:condition, random = ~ 0 + time | grpnbr, data = group_pain)
#summary(group) # this is the one I'm going with I think; I took mean pain for each group, then two levels.
#stargazer(group, type = "text", report = "vcsp", header = F, digits = 2)
```

## Study Design

## Objectives

## Descriptives
|  	| level 	| EDU 	| CBT 	| EET 	| p 	|
|-----------------------------------	|-------	|---------------	|---------------	|---------------	|-------	|
| n 	|  	| 76 	| 75 	| 79 	|  	|
| Therapy site (\%) 	| UM 	| 31 (40.8) 	| 32 (42.7) 	| 38 (48.1) 	| 0.634 	|
|  	| WSU 	| 45 (59.2) 	| 43 (57.3) 	| 41 (51.9) 	|  	|
| Number of sessions (mean (sd)) 	|  	| 6.46 (1.65) 	| 5.93 (2.09) 	| 6.51 (1.85) 	| 0.114 	|
| Age (mean (sd)) 	|  	| 50.28 (12.48) 	| 48.13 (12.54) 	| 48.98 (11.70) 	| 0.552 	|
| Years after diagnosis (mean (sd)) 	|  	| 7.36 (8.19) 	| 8.84 (7.62) 	| 8.86 (8.08) 	| 0.426 	|
| Sex (\%) 	| 1 	| 75 (98.7) 	| 68 (90.7) 	| 73 (92.4) 	| 0.094 	|
|  	| 2 	| 1 ( 1.3) 	| 7 ( 9.3) 	| 6 ( 7.6) 	|  	|
| BMI (mean (sd)) 	|  	| 31.46 (6.37) 	| 30.16 (7.71) 	| 29.16 (6.64) 	| 0.121 	|
| Ethnic (\%) 	| 1 	| 4 ( 5.3) 	| 2 ( 2.7) 	| 2 ( 2.6) 	| 0.601 	|
|  	| 2 	| 72 (94.7) 	| 71 (97.3) 	| 76 (97.4) 	|  	|
| Race (\%) 	| 1 	| 0 ( 0.0) 	| 1 ( 1.3) 	| 1 ( 1.3) 	| 0.361 	|
|  	| 3 	| 1 ( 1.3) 	| 0 ( 0.0) 	| 0 ( 0.0) 	|  	|
|  	| 4 	| 18 (23.7) 	| 15 (20.0) 	| 8 (10.1) 	|  	|
|  	| 5 	| 54 (71.1) 	| 57 (76.0) 	| 68 (86.1) 	|  	|
|  	| 6 	| 3 ( 3.9) 	| 2 ( 2.7) 	| 2 ( 2.5) 	|  	|
| Has health insurance (\%) 	| 0 	| 9 (11.8) 	| 9 (12.0) 	| 4 ( 5.1) 	| 0.244 	|
|  	| 1 	| 67 (88.2) 	| 66 (88.0) 	| 75 (94.9) 	|  	|
| tp total (mean (sd)) 	|  	| 14.75 (2.53) 	| 14.63 (3.00) 	| 14.69 (2.77) 	| 0.963 	|
| acr fmness (mean (sd)) 	|  	| 20.57 (4.73) 	| 20.34 (4.73) 	| 20.96 (4.36) 	| 0.713 	|
| CMSI Total (mean (sd)) 	|  	| 38.76 (12.86) 	| 34.00 (11.72) 	| 37.31 (11.74) 	| 0.050 	|
| Pain at time V2 (mean (sd)) 	|  	| 5.47 (1.74) 	| 5.35 (1.62) 	| 5.34 (1.55) 	| 0.863 	|
| Pain at time V3 (mean (sd)) 	|  	| 5.26 (1.69) 	| 4.74 (1.66) 	| 4.42 (2.04) 	| 0.019 	|
| Pain at time V4 (mean (sd)) 	|  	| 4.86 (1.97) 	| 4.83 (1.69) 	| 4.49 (2.20) 	| 0.462 	|
| Pain difference V2-V4 (mean (sd)) 	|  	| 0.63 (1.92) 	| 0.57 (1.55) 	| 0.80 (1.62) 	| 0.713 	|

## Normality
```{r figure 1, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(long_pain, aes(x = pain)) + geom_density() + facet_wrap(condition ~
    time) + ggtitle("Figure 1: Density plots for pain by treatment and time point")
#ggplot(group_pain, aes(x = mean_pain)) + geom_density() + facet_wrap(condition ~ time) 
```

## Model

## Model results
| Variable            | Value     | Std.Error | p-value |
|---------------------|-----------|-----------|---------|
| timeV3              | -0.223214 | 0.2154765 | 0.3037  |
| timeV4              | -0.576035 | 0.2365247 | 0.0174  |
| timeV2*conditionCBT | -0.097448 | 0.2986305 | 0.7451  |
| timeV3*conditionCBT | -0.557840 | 0.3235078 | 0.0889  |
| timeV4*conditionCBT | -0.120446 | 0.4105077 | 0.7701  |
| timeV2*conditionEET | -0.125255 | 0.2869149 | 0.6637  |
| timeV3*conditionEET | -0.900893 | 0.3108163 | 0.0050  |
| timeV4*conditionEET | -0.411253 | 0.3944031 | 0.3006  |

@@

## Conclusions

>>>>

```{r, message=FALSE, warning=FALSE, results="asis", echo = F}
#group <- lmer(mean_pain ~ time + time:condition + (0 + grpnbr | time), data = group_pain)
#summary(group)
#confint(group, method="boot", nsim=1000)
#group <- lme(mean_pain ~ time + time:condition, random = ~ 1 | grpnbr, data = group_pain)
group <- lme(mean_pain ~ time + time:condition, random = ~ 0 + time | grpnbr, data = group_pain)

#summary(group) # this is the one I'm going with I think; I took mean pain for each group, then two levels.
#stargazer(group, type = "text", report = "vcsp", header = F, digits = 2)
```

## Study Design
- Two-site, three-arm, allegiance-controlled randomized clinical tria
- 230 individual subjects, with 94$\%$ females and 6$\%$ males. 
- Blocks of 18 patients; block broken into three groups of six
- Eight sessions in eight weeks; each session different
- Pain measured at three time points V2, V3, V4
- 36 missing pain values out of 690 pain measurements

## Objectives
1. Test that EET and CBT are more effective in reducing pain levels than the control treatment EDU
2. See if the new EET therapy is more effective than the standard CBT therapy.

## Homogeneity between treatment groups
- Variables were stratified by treatment
- Compared using chi-squared tests if categorical and one-way anova if continuous.

## Descriptive table
|  	| level 	| EDU 	| CBT 	| EET 	| p 	|
|-----------------------------------	|-------	|---------------	|---------------	|---------------	|-------	|
| n 	|  	| 76 	| 75 	| 79 	|  	|
| Therapy site (\%) 	| UM 	| 31 (40.8) 	| 32 (42.7) 	| 38 (48.1) 	| 0.634 	|
|  	| WSU 	| 45 (59.2) 	| 43 (57.3) 	| 41 (51.9) 	|  	|
| Number of sessions (mean (sd)) 	|  	| 6.46 (1.65) 	| 5.93 (2.09) 	| 6.51 (1.85) 	| 0.114 	|
| Age (mean (sd)) 	|  	| 50.28 (12.48) 	| 48.13 (12.54) 	| 48.98 (11.70) 	| 0.552 	|
| Years after diagnosis (mean (sd)) 	|  	| 7.36 (8.19) 	| 8.84 (7.62) 	| 8.86 (8.08) 	| 0.426 	|
| Sex (\%) 	| 1 	| 75 (98.7) 	| 68 (90.7) 	| 73 (92.4) 	| 0.094 	|
|  	| 2 	| 1 ( 1.3) 	| 7 ( 9.3) 	| 6 ( 7.6) 	|  	|
| BMI (mean (sd)) 	|  	| 31.46 (6.37) 	| 30.16 (7.71) 	| 29.16 (6.64) 	| 0.121 	|
| Ethnic (\%) 	| 1 	| 4 ( 5.3) 	| 2 ( 2.7) 	| 2 ( 2.6) 	| 0.601 	|
|  	| 2 	| 72 (94.7) 	| 71 (97.3) 	| 76 (97.4) 	|  	|
| Race (\%) 	| 1 	| 0 ( 0.0) 	| 1 ( 1.3) 	| 1 ( 1.3) 	| 0.361 	|
|  	| 3 	| 1 ( 1.3) 	| 0 ( 0.0) 	| 0 ( 0.0) 	|  	|
|  	| 4 	| 18 (23.7) 	| 15 (20.0) 	| 8 (10.1) 	|  	|
|  	| 5 	| 54 (71.1) 	| 57 (76.0) 	| 68 (86.1) 	|  	|
|  	| 6 	| 3 ( 3.9) 	| 2 ( 2.7) 	| 2 ( 2.5) 	|  	|
| Has health insurance (\%) 	| 0 	| 9 (11.8) 	| 9 (12.0) 	| 4 ( 5.1) 	| 0.244 	|
|  	| 1 	| 67 (88.2) 	| 66 (88.0) 	| 75 (94.9) 	|  	|
| tp total (mean (sd)) 	|  	| 14.75 (2.53) 	| 14.63 (3.00) 	| 14.69 (2.77) 	| 0.963 	|
| acr fmness (mean (sd)) 	|  	| 20.57 (4.73) 	| 20.34 (4.73) 	| 20.96 (4.36) 	| 0.713 	|
| CMSI Total (mean (sd)) 	|  	| 38.76 (12.86) 	| 34.00 (11.72) 	| 37.31 (11.74) 	| 0.050 	|
| Pain at time V2 (mean (sd)) 	|  	| 5.47 (1.74) 	| 5.35 (1.62) 	| 5.34 (1.55) 	| 0.863 	|
| Pain at time V3 (mean (sd)) 	|  	| 5.26 (1.69) 	| 4.74 (1.66) 	| 4.42 (2.04) 	| 0.019 	|
| Pain at time V4 (mean (sd)) 	|  	| 4.86 (1.97) 	| 4.83 (1.69) 	| 4.49 (2.20) 	| 0.462 	|
| Pain difference V2-V4 (mean (sd)) 	|  	| 0.63 (1.92) 	| 0.57 (1.55) 	| 0.80 (1.62) 	| 0.713 	|

## Normality
- For the anovas
```{r figure 1, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(long_pain, aes(x = pain)) + geom_density() + facet_wrap(condition ~
    time) + ggtitle("Figure 1: Density plots for pain by treatment and time point")
#ggplot(group_pain, aes(x = mean_pain)) + geom_density() + facet_wrap(condition ~ time) 
```

## Grouping
- Collapsed on group level with mean pain level within each group at each time point;
- 120 observations (40 groups with 3 measurements each)
- Missing pain values were omitted for mean. 
- Deal with the missing values and missed treatments
- Each group goes through all eight sessions together and is bonded

## Model
- Outcome variable: group pain 
- Covariates: measurement time, interaction between measurement time and treatment type as the covariates
- Measurement time and treatment type treated as categorical
- Only three measurement times, with large, uneven timespans between them
- Assume that baseline pain measurements are equal between the treatment arms
- Random slope

## Model results
| Variable | Value | Std.Error | p-value |
|---------------------|-------|-----------|---------|
| timeV3 | -0.22 | 0.22 | 0.30 |
| timeV4 | -0.58 | 0.24 | 0.02 |
| timeV2*CBT | -0.10 | 0.30 | 0.75 |
| timeV3*CBT | -0.56 | 0.32 | 0.09 |
| timeV4*CBT | -0.12 | 0.41 | 0.77 |
| timeV2*EET | -0.13 | 0.29 | 0.66 |
| timeV3*EET | -0.90 | 0.31 | 0.01 |
| timeV4*EET | -0.41 | 0.40 | 0.30 |
72 degrees freedom

@@

## Conclusions
- Overall, mean pain scores decrease over full time for all three treatments groups
- Mean pain score starts around 5.4 at V2 and ends around 4.7 at V4
- 52 out of the 230 patients (22 percent) of the patients had a 30 percent or greater decrease
- Both EET and CBT treatments decrease pain scores by the end of the treatment (V3), with EET being slightly more effective than CBT. 
- EET and CBT groups retain pain scores at end of study (V4); does not increase
- EDU treatment has no effect at end of treatment (V2), but do have improvement six months after the end of the study (V4)
